Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-19 doc/asplos13/intro.tex
pass on abs,intro

Irregular applications are those whose parallel execution generates tasks whose work, interdependences, or memory accesses are highly sensitive to input.  Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.  Contemporary examples include real-time mining or querying large unstructured data sets in the business, national security, machine learning, data-driven science, and social network computing domains.  For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems.  The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband.  Our vision is to facilitate scaling of irregular applications efficiently on these mass market systems.

What makes our vision challenging is that irregular applications do not readily exploit the features of commodity systems.  This is true even for single node systems.  It is not atypical for any given task's data references to be spread randomly across the entire memory of the system.  Yet, for commodity processors' caches to be effective, computations must exploit spatial locality.  Furthermore, the addresses of references are often determined in the task only just before their values are used.  Unfortunately, mass market prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed.  Because commodity processors' pipeline utilization relies on spatial and temporal locality -- or, at least the ability to predict in advance what is needed -- they stall often when executing irregular applications.

On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.  Irregular applications also present a challenge to mass market network technology, which is designed to transfer blocks of data, not the word-sized references emitted by irregular application tasks.  Injection rate into the networks is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.  While some irregular applications can be restructured to better exploit locality, aggregate requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.

@@

Thus far, we have developed Grappa:  a latency tolerant runtime for scaling irregular applications on x86 clusters.  Grappa leverages as much freely available and commodity infrastructure as possible. We use unmodified Linux for the operating system.  An off the shelf user-mode infiniband device driver stack~\cite{Melonox?} is used for network interfacing.  MPI~\cite{mpi} is used for process setup and tear down.  Finally, GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations.

To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that mitigates the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.  Each of these components will be described in more detail in Section~\ref{sec:grappa}.

@@

\TODO{Add roadmap here}

---

>>>>

Irregular applications are those whose parallel execution generates tasks with work, interdependences, or memory accesses that are highly sensitive to input.  Classic examples of irregular applications include branch and bound optimization, SPICE circuit simulation, contact algorithms in car crash analysis, and network flow.  Contemporary examples include real-time mining or querying large unstructured data sets in the business, national security, machine learning, data-driven science, and social network computing domains.  For these emerging applications, fast response -- given the sheer amount of data -- requires multinode systems.  The most broadly available multinode systems are those built from x86 commodity computing nodes interconnected via ethernet or infiniband.  Our goal is to facilitate scaling of irregular applications efficiently on these mass market systems.

What makes our goal challenging is that irregular applications do not readily exploit the features of commodity systems.  This is true even for single node systems.  It is not atypical for any given task's data references to be spread randomly across the entire memory of the system. Yet, for commodity processors' caches to be effective, computations must exploit spatial locality.  Furthermore, the addresses of references are often determined in the task only just before their values are used.  Unfortunately, commodity prefetching hardware is effective only when addresses are known many cycles before the resident data is consumed.  Because commodity processors' pipeline utilization relies on spatial and temporal locality -- or, at least the ability to predict in advance what is needed -- they stall often when executing irregular applications.

On multinode systems, the challenges presented by low locality are analogous, and exacerbated by the increased latency of going off-node.  Irregular applications also present a challenge to mass market network technology, which is designed to transfer large blocks of data, not the word-sized references emitted by irregular application tasks.  Injection rate into the networks is insufficient to utilize wire bandwidth when blocks are below about two-kilobytes, so any straightforward communication strategy severely under-utilizes the network.  While some irregular applications can be restructured to better exploit locality, aggregate requests to increase message size, and manage the additional challenges of load balance and synchronization across multinode systems, the work to do so is formidable and requires knowledge and skills pertaining to distributed systems far beyond those of most application programmers.

@@

We have developed Grappa:  a latency tolerant runtime for scaling irregular applications on x86 clusters.  Grappa leverages as much freely available and commodity infrastructure as possible. We use unmodified Linux for the operating system.  An off the shelf user-mode infiniband device driver stack~\cite{Melonox?} is used for network interfacing.  MPI~\cite{mpi} is used for process setup and tear down.  GASnet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations. To this commodity hardware and software mix Grappa adds three main software components: (1) a lightweight tasking layer that supports a context switch in as few as \checkme{40ns} and distributed load balancing; (2) a distributed shared memory layer that supports customary operations such as read and write as well as remote operations such as fetch-and-add~\cite{fetchandadd}; and (3) a message aggregation layer that combines short messages to mitigate the aforementioned problem that commodity networks are designed to achieve peak bandwidth only on large packet sizes, yet irregular applications tend to fetch only a handful of bytes at a time.  Each of these components will be described in more detail in Section~\ref{sec:grappa}.

@@

\TODO{Add roadmap here. Luis: do we really want a roadmap here? We have enough about what we have done I think, so we might not need to sell things we plan to do.}
